---
title: "Synthetic Data"
format:
  html:
    toc: false
    number-sections: true
    highlight-style: github
---

Create a synthetic dataset for a clinical trial assessing second-line immune checkpoint inhibitor therapy (pembrolizumab) in the treatment of advanced urothelial carcinoma.

Data was taken from a randomized controlled trial (RCT) by Bellmunt et al., 2017. 

The data summary used to generate the synthetic data is as follows:
Pembrolizumab group:
Age (mean) in years = 67
Male sex = 74.1%
Current or former smoker = 61.3%
Site of primary tumor in bladder or urethra = 85.9%
Median overall survival = 10.3 months
Median progression free survival = 2.1 months
Adverse event leading to discontinuation = 5.6%

Chemotherapy group:
Age (mean) in years = 68
Male sex = 74.3%
Current or former smoker = 69.1%
Site of primary tumor in bladder or urethra = 86.3%
Median overall survival = 7.4 months
Median progression free survival = 3.3 months
Adverse event leading to discontinuation = 11%

Create synthetic data using this data:
```{r}
# Set random seed for reproducibility
set.seed(123)

# Function to simulate survival times
simulate_survival <- function(n, median_survival) {
  lambda <- log(2) / median_survival
  rexp(n, rate = lambda)
}

# Number of synthetic data points for each group
n_pembrolizumab <- 1000
n_chemotherapy <- 1000

# Pembrolizumab group
pembrolizumab_data <- data.frame(
  Group = "Pembrolizumab",
  Age = rnorm(n_pembrolizumab, mean = 67, sd = 10),
  Male = rbinom(n_pembrolizumab, 1, 0.741),
  Smoker = rbinom(n_pembrolizumab, 1, 0.613),
  Primary_Tumor_Bladder_Urethra = rbinom(n_pembrolizumab, 1, 0.859),
  Overall_Survival = simulate_survival(n_pembrolizumab, 10.3),
  Progression_Free_Survival = simulate_survival(n_pembrolizumab, 2.1),
  Adverse_Event_Discontinuation = rbinom(n_pembrolizumab, 1, 0.056)
)

# Chemotherapy group
chemotherapy_data <- data.frame(
  Group = "Chemotherapy",
  Age = rnorm(n_chemotherapy, mean = 68, sd = 10),
  Male = rbinom(n_chemotherapy, 1, 0.743),
  Smoker = rbinom(n_chemotherapy, 1, 0.691),
  Primary_Tumor_Bladder_Urethra = rbinom(n_chemotherapy, 1, 0.863),
  Overall_Survival = simulate_survival(n_chemotherapy, 7.4),
  Progression_Free_Survival = simulate_survival(n_chemotherapy, 3.3),
  Adverse_Event_Discontinuation = rbinom(n_chemotherapy, 1, 0.11)
)

# Combine the two groups
synthetic_data <- rbind(pembrolizumab_data, chemotherapy_data)

# View summary of the synthetic data
summary(synthetic_data)

```

Now create plot to look at overall survival vs group
```{r}
library(ggplot2)
ggplot(synthetic_data, mapping=aes(x=Group, y=Overall_Survival)) + geom_point()
```

As expected, the overall survival is greater in pembrolizumab group than in chemotherapy group.

Now a plot to look at progression free survival vs group
```{r}
ggplot(synthetic_data, mapping=aes(x=Group, y=Progression_Free_Survival)) + geom_point()
```

As expected, progression free survival is greater in chemotherapy group.

Now fit a model to predict overall survival based on age, sex, smoking status, site of primary tumor, group, and adverse events

```{r}
survival_fit <- lm(
  Overall_Survival ~ Group + Age + as.factor(Male) + 
    as.factor(Smoker) + as.factor(Primary_Tumor_Bladder_Urethra) + 
    as.factor(Adverse_Event_Discontinuation),
  data = synthetic_data
)

summary(survival_fit)
```

We can see that overall survival depends on the treatment group and age, but not on other factors such as sex, smoking status, site of primary tumor, and adverse events.
The R-squared is 0.02259, suggesting that there are other factors that may influence overall survival.